%=================================================================
% HOW TO USE THIS FILE
%=================================================================
% This file contains the manuscript CONTENT only. It is NOT a
% compilable MDPI submission on its own: MDPI requires their own
% class file (mdpi.cls) and a "Definitions/" folder (logos,
% bibliography styles, journal-specific settings), which are NOT
% reproduced here (they are MDPI's own files, not something to
% copy from a third party -- download them fresh, official source
% only, right before finalizing the submission, since MDPI updates
% the template periodically).
%
% Steps to assemble the real submission:
%   1. Go to https://www.mdpi.com/authors/latex and download the
%      current template ZIP (choose the citation style your target
%      journal uses -- ACS/numbered is the default for most
%      MDPI science journals; APA/author-date for a shorter list,
%      Climate uses NUMBERED citations (\cite), not author-date -- see
%      note near \bibliography at the end of this file.
%   2. Unzip it; you will get template.tex + a Definitions/ folder.
%   3. Replace the placeholder journal name below
%      (\documentclass[JOURNALNAME,article,submit,moreauthors]{Definitions/mdpi})
%      with your target journal's exact string, copied from the
%      comment block in MDPI's own template.tex (case-sensitive).
%   4. Copy everything from \Title{...} down to \end{document} in
%      THIS file into MDPI's template.tex, replacing their
%      placeholder content.
%   5. Copy references.bib into the same project folder.
%=================================================================

\documentclass[climate,article,submit,moreauthors]{Definitions/mdpi}
% <-- Confirm 'climate' matches exactly the string in the current official
%     template.tex comment block before compiling (case-sensitive). Chosen
%     over Entropy and over Atmosphere (which already hosts the companion
%     MF-DFA manuscript, so a different journal was preferred for this paper).

%=================================================================
% MDPI internal commands -- fill in once the target journal and
% submission stage are decided; leave as-is for now.
%=================================================================
\firstpage{1}
\makeatletter
\setcounter{page}{\@firstpage}
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2026}
\copyrightyear{2026}
\datereceived{}
\daterevised{}
\dateaccepted{}
\datepublished{}

%=================================================================
% TITLE -- placeholder, to be written last per the agreed workflow
%=================================================================
\Title{[TITLE PENDING]}

%=================================================================
% AUTHORS
%=================================================================
\TitleCitation{[Short title for citation -- PENDING]}
\newcommand{\orcidauthorA}{0000-0000-0000-0000} % <-- fill in ORCID if available
\Author{Rafael Magallanes-Quintanar $^{1,}$*}
\AuthorNames{Rafael Magallanes-Quintanar}
\AuthorCitation{Magallanes-Quintanar, R.}

\address{%
$^{1}$ \quad Unidad Acad\'emica de Ingenier\'ia El\'ectrica, Programa de Ingenier\'ia en Computaci\'on, Universidad Aut\'onoma de Zacatecas, Zacatecas, Mexico
}

\corres{Correspondence: [email] }

%=================================================================
% ABSTRACT AND KEYWORDS -- placeholders, written last
%=================================================================
\abstract{[ABSTRACT PENDING -- to be written after Discussion is finalized.]}

\keyword{recurrence quantification analysis; cross recurrence plot; drought;
Standardized Precipitation Index; nonlinear time series analysis; synchronization;
Zacatecas; Mexico}

%=================================================================
\begin{document}

% ================================================================
\section{Introduction}
\label{sec:introduction}
[PENDING -- section to be drafted after Discussion.]

% ================================================================
\section{Materials and Methods}
\label{sec:methods}

\subsection{Study area and data}
\label{sec:methods-data}

The study area comprises the state of Zacatecas, north-central Mexico, partitioned
into four physiographic regions (Semi-arid, Highlands, Mountains, Canyons) following
the official regionalization established in the authors' related work on Multifractal
Detrended Fluctuation Analysis (MF-DFA) of regional drought series
\cite{MFDFAmanuscript}. This regionalization is based on 29--31 CONAGUA
meteorological stations grouped by physiographic and hydroclimatic criteria, and was
independently validated at the regional-aggregation level via bootstrap analysis of
multifractal features in that companion study.

The 12-month Standardized Precipitation Index (SPI-12) was computed for each region
from monthly precipitation records spanning \textbf{December 1965 to December 2024
(709 months)}. Station-level series were quality-controlled, gap-filled, and
aggregated to the regional level using the same pipeline underlying the MF-DFA
manuscript; full details of station selection, gamma-distribution fitting, and
imputation are provided therein and are not repeated here. The four resulting
regional SPI-12 series constitute the sole input to the analyses described below.

\subsection{Phase-space reconstruction and embedding parameter selection}
\label{sec:methods-embedding}

Recurrence-based methods require reconstructing the system's phase space from a
scalar time series via time-delay embedding \cite{Takens1981}. For a series
$x(t)$, an embedded state vector is defined as

\begin{equation}
\vec{x}_i = \big(x(i),\, x(i+\tau),\, x(i+2\tau),\, \dots,\, x(i+(m-1)\tau)\big)
\label{eq:embedding}
\end{equation}

where $\tau$ is the delay and $m$ is the embedding dimension.

For each of the four regional series, $\tau$ was estimated as the first local
minimum of the average mutual information (AMI) function, computed over lags
1--24 months using a 16-bin two-dimensional histogram estimator. The embedding
dimension $m$ was estimated via the false nearest neighbors (FNN) method
\cite{Kennel1992}, selecting the smallest $m$ at which the percentage of false
neighbors fell below 5\%, with standard thresholds $R_{tol}=15$, $A_{tol}=2$.
Resulting parameters are reported in Section~\ref{sec:results-embedding}
(Table~\ref{tab:embedding}).

Because the sliding-window synchronization analysis
(Section~\ref{sec:methods-sliding}) operates on much shorter sub-series
(12--36 months) than the full 709-month series, applying the same, comparatively
high-dimensional ($m\approx4$--5, $\tau\approx9$--11 months) embedding to short
windows was found to leave zero usable embedded vectors in many windows
($(m-1)\tau$ alone can consume 27--40 months). For that analysis only, a minimal
fixed embedding ($m=2$, $\tau=1$) was used instead --
see Section~\ref{sec:methods-sliding}.

\subsection{Recurrence plots and RQA measures}
\label{sec:methods-rqa}

For each region, a recurrence plot (RP) was constructed from the embedded
trajectory:

\begin{equation}
R_{i,j} = \Theta(\varepsilon - \lVert \vec{x}_i - \vec{x}_j \rVert)
\label{eq:rp}
\end{equation}

where $\Theta$ is the Heaviside function, $\lVert \cdot \rVert$ is the Euclidean
norm, and $\varepsilon$ is a distance threshold. Rather than fixing $\varepsilon$
to an arbitrary value (e.g., a fixed multiple of the signal's standard
deviation), $\varepsilon$ was selected independently for each region as the
empirical percentile of the pairwise-distance distribution corresponding to a
\textbf{target recurrence rate of RR = 5\%}, following standard recommendations
\cite{Marwan2007} to keep RQA measures comparable across series with different
amplitude scales.

\textbf{Theiler window correction.} Points adjacent in time are trivially close
in phase space whenever the underlying signal is smooth -- which SPI-12, as a
12-month moving average, inherently is. Including such trivial,
autocorrelation-driven matches inflates diagonal-line-based measures (see below)
without reflecting genuine recurrence to a distinct past or future state
\cite{Theiler1986}. To correct for this, a band of half-width $W$ around the
main diagonal ($|i-j| < W$) was excluded from both the $\varepsilon$-calibration
step and the final recurrence matrix. We set $W = \tau$ per region, using the
decorrelation timescale already estimated in
Section~\ref{sec:methods-embedding} as a principled, data-driven choice rather
than an arbitrary one. All RQA measures reported in this manuscript use the
Theiler-corrected recurrence matrices.

From the corrected recurrence matrix, the following standard RQA measures
\cite{Marwan2007} were computed from the distributions of diagonal and
vertical line lengths (minimum line length = 2 for both):

\begin{itemize}
  \item \textbf{RR} (recurrence rate): density of recurrence points,
  $RR = \frac{1}{N^*}\sum_{i,j} R_{i,j}$, where $N^*$ is the number of $(i,j)$
  pairs outside the Theiler-excluded band.
  \item \textbf{DET} (determinism): fraction of recurrence points forming
  diagonal lines of length $\geq L_{min}$.
  \item \textbf{L}: mean diagonal line length.
  \item \textbf{Lmax}: length of the longest diagonal line.
  \item \textbf{ENTR}: Shannon entropy of the diagonal line-length distribution.
  \item \textbf{LAM} (laminarity): fraction of recurrence points forming
  vertical lines of length $\geq V_{min}$ -- an operational measure of states
  "trapped" in a regime.
  \item \textbf{TT} (trapping time): mean vertical line length -- the typical
  duration of a regime before transitioning.
\end{itemize}

\subsection{Cross-Recurrence and the F-synchronization measure}
\label{sec:methods-crp}

Following the approach of Semeraro et al. \cite{Semeraro2020}, pairwise synchronization between regions was
assessed via the Cross Recurrence Plot (CRP):

\begin{equation}
C_{i,j} = \Theta(\varepsilon - \lVert \vec{x}_i - \vec{y}_j \rVert)
\label{eq:crp}
\end{equation}

where $\vec{x}$ and $\vec{y}$ are the embedded trajectories of two different
regional series, with $\varepsilon$ again calibrated to a target RR = 5\% for
the pair. Synchronization was quantified with the \textbf{F-recurrence measure},
the density of points along the CRP's main diagonal (the line of
synchronization):

\begin{equation}
F = \frac{1}{N}\sum_i C_{i,i}
\label{eq:frecurrence}
\end{equation}

$F=1$ indicates complete synchronization; $F \to 0$ indicates desynchronization.
Unlike the single-region RQA measures in Section~\ref{sec:methods-rqa}, $F$ is
computed only along matched time indices $(i,i)$, not across a range of
offsets, and is therefore not subject to the Theiler-window artifact described
above; no correction was applied to this measure.

\subsection{Sliding-window synchronization around documented drought events}
\label{sec:methods-sliding}

To evaluate whether inter-regional synchronization changes around severe
drought events, $F$ was recomputed in 12-month sliding windows (3-month step)
within 36-month periods immediately before and after each of five
independently documented drought years (1996, 2011, 2012, 2021, 2023; sources
and event-specific context in Table~\ref{tab:droughtevents}). Because these
short windows are incompatible with the full-series embedding parameters
(Section~\ref{sec:methods-embedding}), a fixed minimal embedding ($m=2$,
$\tau=1$) was used for this analysis only. Windows yielding fewer than 5
embedded vectors were skipped (this occurred systematically for the 2023
event, for which the series' end date, December 2024, allows only 12 of the
intended 36 post-event months).

For each region pair and event with $\geq 2$ valid windows on both sides, the
pre/post distributions of $F$ were compared with Levene's test (variance
homogeneity) and Welch's $t$-test (difference of means, unequal variances).
Drought years were selected from independently documented sources (national
CONAGUA/SMN records and, where available, Zacatecas-specific peer-reviewed
literature and direct grower testimony; see Table~\ref{tab:droughtevents}),
not chosen post hoc from the data.

\subsection{Multiple-comparison correction}
\label{sec:methods-mcc}

The sliding-window analysis (Section~\ref{sec:methods-sliding}) produces one
test per region pair per event (up to 6 pairs $\times$ 5 events = 30 tests,
reduced to 24 valid tests after excluding windows with insufficient data). At
$\alpha=0.05$, roughly one false positive is expected by chance alone across
this many tests. Both a conservative Bonferroni correction and the less
conservative Benjamini-Hochberg false discovery rate (FDR) procedure were
applied to the resulting $p$-values. One region pair-event
(Semi-arid--Highlands, 2021) was treated as a pre-specified, confirmatory test
rather than an exploratory one, since independent field evidence (grower
testimony) identified 2021 as a Highlands-concentrated event \textit{prior to}
running the analysis; for this test, the FDR-corrected result (not Bonferroni)
is considered the appropriate significance criterion (see
Section~\ref{sec:results-sliding} and Discussion).

\subsection{Software and reproducibility}
\label{sec:methods-software}

All analyses were implemented in Python (NumPy, SciPy, pandas) in a Google
Colab notebook, using a custom implementation of RQA rather than third-party
RQA packages, for full transparency and auditability of every computational
step (embedding, threshold calibration, Theiler-window exclusion, and
line-length extraction). The complete notebook, including the data-loading,
verification, and all diagnostic checks described in
Section~\ref{sec:results}, is provided as supplementary material and archived
on GitHub [\textbf{PENDING: repository URL / DOI}].

% ================================================================
\section{Results}
\label{sec:results}

\subsection{Embedding parameter selection}
\label{sec:results-embedding}

For each of the four regional series of SPI-12 (Semi-arid, Highlands,
Mountains, Canyons; 709 months, December 1965--December 2024), the embedding
delay ($\tau$) and dimension ($m$) were estimated as described in
Section~\ref{sec:methods-embedding} (Table~\ref{tab:embedding}).

\begin{table}[H]
\caption{Embedding parameters per region.}
\label{tab:embedding}
\newcolumntype{C}{>{\centering\arraybackslash}X}
\begin{tabularx}{\linewidth}{lCC}
\toprule
\textbf{Region} & \boldmath{$\tau$} & \boldmath{$m$} \\
\midrule
Semi-arid & 9  & 5 \\
Highlands & 11 & 4 \\
Mountains & 10 & 5 \\
Canyons   & 11 & 5 \\
\bottomrule
\end{tabularx}
\end{table}

The obtained $\tau$ values (9--11 months) are consistent with the seasonal
periodicity inherent to SPI-12 as a 12-month moving window. The embedding
dimension resulted between 4 and 5 across the four regions, with Highlands as
the only region with $m=4$ instead of $m=5$.

\subsection{Regional RQA measures}
\label{sec:results-rqa}

With the recurrence threshold fixed to achieve a constant recurrence rate
(RR$\approx$0.05) across the four regions, and after excluding a Theiler
window of width $W=\tau$ (per region) around the main diagonal to remove
trivial autocorrelation-driven recurrence \cite{Theiler1986,Marwan2007},
standard RQA measures were computed (Table~\ref{tab:rqa}).

\begin{table}[H]
\caption{RQA measures per region (Theiler-corrected).}
\label{tab:rqa}
\newcolumntype{C}{>{\centering\arraybackslash}X}
\begin{tabularx}{\linewidth}{lCCCCCCC}
\toprule
\textbf{Region} & \textbf{RR} & \textbf{DET} & \textbf{L} & \textbf{Lmax} & \textbf{ENTR} & \textbf{LAM} & \textbf{TT} \\
\midrule
Semi-arid & 0.050 & 0.376 & 3.346 & 21 & 1.582 & 0.854 & 4.220 \\
Highlands & 0.050 & 0.380 & 3.328 & 21 & 1.582 & 0.864 & 4.458 \\
Mountains & 0.050 & 0.374 & 3.209 & 21 & 1.515 & 0.849 & 4.188 \\
Canyons   & 0.050 & 0.382 & 3.708 & 30 & 1.751 & 0.863 & 4.895 \\
\bottomrule
\end{tabularx}
\end{table}

\textit{Methodological note (Theiler window):} an initial run without
excluding the near-diagonal band produced Lmax values of 143--343 months,
driven in the most extreme case (Mountains) entirely by an offset-1 diagonal
-- i.e., consecutive months trivially resembling each other due to the
smoothness of SPI-12 as a 12-month moving average, not genuine recurrence to
a distinct past state. Following the standard correction of
Theiler \cite{Theiler1986}, a band of width $W=\tau$ was excluded from all RQA
computations; Table~\ref{tab:rqa} reflects the corrected values used
throughout this manuscript.

Canyons presents the highest values of DET, L, Lmax, ENTR, LAM, and TT among
the four regions -- this ranking is unchanged by the Theiler-window
correction, indicating it is a robust feature of the data rather than an
artifact of the near-diagonal band. Differences between regions remain modest
in absolute terms (e.g., LAM between 0.849 and 0.864), consistent with shared
regional climatic forcings across all four regions.

\textit{Verification of Lmax (Mountains):} the longest diagonal line
(length = 21 months) after correction links two temporally distant wet
episodes -- July 1985--March 1987 and March 2002--November 2003, offset by
200 months ($\approx$16.7 years) -- both with positive mean SPI-12 (0.30 and
0.36, respectively) and healthy within-segment variability (std 0.63 and
0.39; 21/21 unique values in both segments, ruling out a data artifact). A
direct Pearson correlation between the two segments (aligned by relative
month) confirms a moderate-to-strong positive relationship ($r=0.53$),
consistent with genuine recurrence between two similar wet regimes rather
than a methodological artifact.

\subsection{Synchronization between regions (Cross-Recurrence)}
\label{sec:results-crp}

The F-synchronization measure (density of the identity line in the Cross
Recurrence Plot; \cite{Semeraro2020}) was computed for the six pairwise
region combinations (Table~\ref{tab:frecurrence}).

\begin{table}[H]
\caption{Pairwise F-recurrence (synchronization) between regions.}
\label{tab:frecurrence}
\newcolumntype{C}{>{\centering\arraybackslash}X}
\begin{tabularx}{0.6\linewidth}{lC}
\toprule
\textbf{Region pair} & \textbf{F} \\
\midrule
Semi-arid -- Highlands & \textbf{0.639} \\
Highlands -- Mountains & 0.573 \\
Semi-arid -- Mountains & 0.461 \\
Mountains -- Canyons   & 0.460 \\
Highlands -- Canyons   & 0.352 \\
Semi-arid -- Canyons   & 0.287 \\
\bottomrule
\end{tabularx}
\end{table}

Semi-arid and Highlands constitute, by a clear margin, the most synchronized
pair -- consistent with their geographic contiguity in the inter-range basin
between the Sierra Madre Oriental and Occidental. Semi-arid and Canyons
constitute the least synchronized pair, consistent with distinct orographic
precipitation regimes between the two regions.

\subsection{Synchronization before/after documented drought events}
\label{sec:results-sliding}

The change in the $F$ measure was evaluated before and after five
independently documented drought events (Table~\ref{tab:droughtevents}), for
the six region-pair combinations. Of the resulting 24 tests (2023 was
excluded due to insufficient post-event data within the series range),
correction for multiple comparisons was applied (Table~\ref{tab:sliding}).

\begin{table}[H]
\caption{Region pairs with statistically significant synchronization change around a drought event (after multiple-comparison correction).}
\label{tab:sliding}
\newcolumntype{C}{>{\centering\arraybackslash}X}
\begin{tabularx}{\linewidth}{lCCCCCC}
\toprule
\textbf{Pair} & \textbf{Event} & \textbf{F before} & \textbf{F after} & \boldmath{$p$} \textbf{(raw)} & \boldmath{$p$} \textbf{(Bonf.)} & \textbf{Sig. FDR} \\
\midrule
Semi-arid--Highlands & 2012 & 0.020 & 0.323 & $2.27\times10^{-8}$ & $5.45\times10^{-7}$ & \checkmark \\
Highlands--Canyons   & 2021 & 0.010 & 0.253 & $2.45\times10^{-3}$ & $5.89\times10^{-2}$ & \checkmark \\
Semi-arid--Highlands & 2021 & 0.313 & 0.091 & $8.10\times10^{-3}$ & $1.94\times10^{-1}$ & \checkmark \\
\bottomrule
\end{tabularx}
\end{table}

The three remaining pair-events with uncorrected $p<0.05$
(Semi-arid--Canyons 2012, $p=0.025$; Highlands--Canyons 2011, $p=0.034$;
Mountains--Canyons 2011, $p=0.036$) did not survive any correction and are
reported as inconclusive trends, not as findings. The 1996 event did not
produce significant changes in any region pair ($p>0.05$ across all five
pairs evaluated).

The two events documented as being of systemic/national scale within the
period with sufficient data (2011--2012) are associated with a marked
increase in synchronization in the geographically best-connected pair
(Semi-arid--Highlands). The 2021 event, documented via grower testimony as
regionally concentrated in Highlands municipalities (Sombrerete, Valparaíso,
Río Grande), produced a divergent response across pairs: synchronization in
Highlands--Canyons but desynchronization in Semi-arid--Highlands -- the pair
that under normal conditions is the most strongly coupled
(Section~\ref{sec:results-crp}).

\begin{table}[H]
\caption{Documented drought events used in the sliding-window analysis.}
\label{tab:droughtevents}
\newcolumntype{C}{>{\centering\arraybackslash}X}
\begin{tabularx}{\linewidth}{lXl}
\toprule
\textbf{Event} & \textbf{Documentation} & \textbf{Scale} \\
\midrule
1996 & Worst drought of the 1990s nationally; paralyzed agricultural exports; mass livestock sacrifice \cite{TVAztecaSequias} & Systemic/national \\
2011 & Record affected area since satellite records began; 95.12\% of national territory affected, driven by La Ni\~na \cite{SMNsequia2026} & Systemic/national, continental \\
2011--2012 & Worst drought in 70 years in Zacatecas; $\sim$13-month dry spell; $>$75,000 head of livestock lost \cite{Scielo2016Sequia,Scielo2018Sequia,AliatConexxion2012} & Zacatecas, systemic \\
2021--2022 & $>$85\% of country critical (Apr. 2021); urban water crisis in northern states by 2022 \cite{SMNsequia2026} & Continental/national \\
2021 & Critical ranching year in Sombrerete, Valparaíso, R\'io Grande (Highlands); direct grower testimony & Zacatecas, regionally concentrated \\
2023--2024 & 76\% of national territory D0--D4; reservoirs at historic lows \cite{Informador2024Sequia,WikipediaEscasezHidrica,Ganaderia2024Sequia} & Systemic/national \\
2023 & 50 of 58 Zacatecas municipalities in extreme drought (CONAGUA); not statistically testable, data ends Dec. 2024 & Zacatecas, systemic \\
\bottomrule
\end{tabularx}
\end{table}

% ================================================================
\section{Discussion}
\label{sec:discussion}
[PENDING]

% ================================================================
\section{Conclusions}
\label{sec:conclusions}
[PENDING]

% ================================================================
% BACK MATTER -- MDPI requires these sections even if brief.
% Fill in before submission.
% ================================================================

\vspace{6pt}
\authorcontributions{[PENDING -- CRediT taxonomy statement.]}
\funding{[PENDING]}
\institutionalreview{Not applicable.}
\informedconsent{Not applicable.}
\dataavailability{The regional SPI-12 series and the complete analysis
notebook are available at [\textbf{PENDING: GitHub repository URL}].}
\acknowledgments{[PENDING]}
\conflictsofinterest{The authors declare no conflict of interest.}

%=================================================================
% References -- Climate uses NUMBERED citation style (\cite), already
% applied throughout this file. Re-verify against the current
% instructions-for-authors page before submission, in case MDPI has
% changed this journal's convention.
%=================================================================
\reftitle{References}
\bibliography{references}

\end{document}
